What is an Ancestral Recombination Graph?

### 2022-02-11 paper.tex

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{tikz}

\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

@@

This original formulation which does not keep track of the distribution
of ancestral material among lineages is the so-called ``Big'' ARG.
The ``little'' ARG is the subset of the Big ARG traversed by
Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG]

@@

\begin{figure}
\centering
\begin{tikzpicture}[xscale=0.8, yscale=0.3, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] {};
\node (s1) [right=of s0,greynode] {};
\node (s2) [right=of s1,greynode] {};
\node [greynode] (s3) at ($(s1) + (0,1)$) {};
\node [greynode] (s4) at ($0.5*(s1) + 0.5*(s2) + (0,2)$) {};
\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
\node [greynode] (s6) at ($(s1) + (0,5)$) {};

\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x = 0.3$};
\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\foreach \child/\parent in {s6/s5, s6/s4, s5/s0, s4/s2}
	\draw (\child) -| (\parent);

\draw (s5) -| ($0.5*(s3) + 0.5*(s3-|s5)$) |- (s3);
\draw (s4) -| ($0.5*(s3) + 0.5*(s3-|s4)$) |- (s3);

\end{tikzpicture}
\caption{\label{fig-arg} \textcolor{red}{A picture of a standard ARG for us
to work with for now. We should replace it with something better at some
point.}
 }
\end{figure}

@@

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage format.

@@

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

@@

This original formulation which does not keep track of the distribution
of ancestral material among lineages is the so-called ``Big'' ARG.
The ``little'' ARG is the subset of the Big ARG traversed by
Hudson's algorithm (\citet{wiuf1999recombination} refer to this as HUD).
[Review of other work done on the ARG stochastic process here.
Discuss \citet{wiuf1999ancestry}. Wiuf and Hein's classical pair of papers
clarified the relationship between Hudson's algorithm and the ARG.
Explain that ARG simulation is inherently less efficient than Hudson's
algorithm because we must track all lineages, even those that correspond
to sections that have coalesced, or we won't have a complete ARG.
This might be tricky to communicated without a figure, but is an
esoteric point perhaps not worth spending a figure on.]

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (3, 1) {};
\node (s4) [greynode] at (1, 2) {};
\node (s5) [greynode] at (5, 3) {};
\node (s6) [greynode] at (3, 4) {};

\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\draw (s0) |- (s4);
\draw (s4) -- (2,2) |- (s3);
\draw (s4) |- (s6);
\draw (s3) -- (4,1) |- (s5);
\draw (s2) |- (s5);
\draw (s5) |- (s6);

\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
\begin{tabular}{c|c}
\multicolumn{2}{c}{Breakpoint}\\
\hline
0 & $\infty$ \\
1 & $\infty$ \\
2 & $\infty$ \\
3 & $x$ \\
4 & $\infty$ \\
5 & $\infty$ \\
6 & $\infty$ \\
\end{tabular}};

\node [nodelabel,anchor=north west] at ($(-2.5,5)$) {
\begin{tabular}{l}
Edges\\
\hline
$(0, 4)$ \\
$(1, 3)$ \\
$(2, 5)$ \\
$(3, 4)$ \\
$(3, 5)$ \\
$(4, 6)$ \\
$(5, 6)$ \\
\end{tabular}};

\end{tikzpicture}
&
\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (2, 1) {};
\node (s4) [greynode] at (4, 1) {};
\node (s5) [greynode] at (1, 2) {};
\node (s6) [greynode] at (5, 3) {};
\node (s7) [greynode] at (3, 4) {};

\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) |- (s3);
\draw (s1) |- (s4) |- (s6);
\draw (s3) |- (s5);
\draw (s0) |- (s5);
\draw (s2) |- (s6);
\draw (s5) |- (s7);
\draw (s6) |- (s7);

\node [nodelabel,anchor=north west] at ($(-5,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(0, 5)$ & $(0, L]$ \\
$(1, 3)$ & $(0, x]$ \\
$(1, 4)$ & $(x, L]$ \\
$(2, 6)$ & $(0, L]$ \\
$(3, 5)$ & $(0, L]$ \\
$(4, 6)$ & $(0, L]$ \\
$(5, 7)$ & $(0, L]$ \\
$(6, 7)$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process
associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format.
The concrete details of how an ARG should be stored and processed
are not discussed in the literature, % weak, fix.
but the following is a reasonable approach that mirrors the implementations
in methods such as [x, y z].
An ARG is composed of a set of $N$ nodes, where each node is an
integer $0 \leq u < N$, and edges $(u, v)$ define the relatships between
those nodes.
Each node corresponds to an \emph{event}, and is either a common ancestor
event (in which case it has two children and one parent)
or a recombination event (when it has one child and two parents).
Each recombination event is associated with a breakpoint $0 < x < L$,
which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
node and $f(u) = 0$ otherwise.

@@

\item Comparison metrics are a problem. All evaluations has essentially
looked at either some form of pairwise comparison of trees along the genome
using tree metrics, or looked at the distributions of TMRCAs. No
one has succeeded in performing any comparisons of ARGs beyond
regarding them as a sequence of independent trees. If we genuinely believe
that ARGs are more than this, then we need to devise some metrics that
consider more than the local tree structure.
